### 2024-05-30 journal/response.txt

> line 558, right column. The sentence "the block's element field is
> $e$ to indicate it represents Enqueue(e) operation..." is
> confusing. It's not immediately clear why element $e$ indicates
> that the operation is Enqueue. Probably formulate it "the block's
> element field $e\neq null$ indicates..."

This paragraph has been rewritten.

> lines 594 - 608 (lines 24 - 37 in the code). Add a more detailed
> comments that explain the Refresh operation.

Done.  Quite a few new comments have been added throughout the pseudocode.

@@

We found it easier to combine Theorem 22 and Corollary 23 and their
proofs (now stated together as Theorem 22).

> line 1304, right column. "...will also keep the running time of our
> operations polylogarithmic *on $p$ and $q$*"

Done.

> Reviewer 2
>
> Then, the paper presents a modification of the first algorithm that uses bounded memory, more specifically polynomial on the number of processes and the size of the queue in the linearization, and with polylogarithmic amortized step complexity. As one would naturally expect, this version adds a sort of garbage collector mechanism that removes blocks of memory that are not needed anymore. I found this part of the paper harder to understand. I did not understand the need of using red-black trees, which I think complicates everything. I have no reason to think the implementation is wrong, all looks convincing to me, but I wonder if there is a way to simplify the presentation.

For the space-bounded version, we replace an infinite array by a red-black tree (using the array index as the key).  This allows us to append elements, access elements, and split off the obsolete elements, all in logarithmic time.  We realize that this additional data structure complicates the algorithm, but we were unfortunately unable to come up with a simpler way to do it.  We hoped that presenting the simpler, space-unbounded version first would make it easier to understand the space-bounded version.  We did our best to explain the transformation from the arrays to the RBTs in the first page of Section 6.

> Specific comments:
> 
> - ln 37, left. Missing blank space before ‘and’.

@@

Fixed.

> - I find the related work section complete. However, I wonder if it is worth citing a few works about memory reclamation, due to your bounded memory implementation.

Although our space-bounded algorithm unlinks objects that are no longer needed from our data structure, we consider the problem of reclaiming that memory to be an orthogonal issue.  There is a rather large body of literature on memory reclamation for concurrent data structures, so we have added (at line 110 of the introduction) a reference to a very recent paper that includes a large bibliography of work in this area, to help the reader find relevant literature on it.

@@

> - I do not understand why you do not cite the preliminary version in podc 2023.

This omission has been corrected.

> - ln 235, left. Repeated ‘words’.
> 
> - For completeness, I suggest to add a preliminaries section, with a reminder of the main definitions such as linearizability, step complexity, and so on. 
> 
> - ln 417, right. Are these direct blocks?
> 
> - ln 492-498. I would recall here what q stands for. Actually, I wonder if you could you use a letter that is more descriptive. 
> 
> - Sec 3.4, and later. At many places operations are written in different font styles. Please follow a uniform format. 
> 
> - ln 845-848, right. I got confused here. Lemma 4 is only about the indexes, isn't it? 
> 
> - ln 1084-1088, right. Here do you mean that L is a valid sequential execution of the queue.
> 
> - Lemma 15. After reading the proof I understood that legal linearization ordering only  means respecting real-time order. This is a concrete example why I think a preliminaries section might help.
> 
> - ln 1192-1993, right. Please formalize the notion of time in an execution,
> 
> - Theorem 22. What exactly is the size of the queue?
> 
> - Sec 6. Why do you need read-back trees? How is that they facilitate the task of keeping memory bounded? Why the first algorithm cannot be modified to remove bocks, let us say using some sort of epoch-based memory reclamation scheme?
> 
> - ln 1279-1283, left. Why? A whole path might need to be copied, isn’t it?
> 
> - ln 1442. Why not Help(), in accordance to Dequeue(), for example?
> 
> - Sec 7. Your algorithms are adaptive on the size of the queue but not on the number of processes. I wonder if an interesting future direction would be to make your algorithms fully adaptive.
>

### 2024-05-30 journal/introduction.tex
fix typo: add misssing space
working on reviewer 2's comments

There has been a great deal of research in the past several decades on the design of shared queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g.,\cite{BBFRSW21}),
% ouroboros paper by Winter et al 2020 might be a source for queues used for mem management (in context of GPUs)
%packet processing \cite{DPDK}, 
synchronization \cite{KAE23},%\here{is this a good citation for this?}
and sharing resources or tasks.
We focus on shared queues that are \emph{linearizable} \cite{HW90}, meaning that operations
appear to take place atomically, and \emph{lock-free}, meaning that some operation on the queue
is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.
We study the \emph{amortized step complexity} per operation, which is measured by taking the maximum, over all possible
executions, of the number of steps in the execution divided by the total number of enqueue and dequeue operations in the execution.

@@

Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed by Ruppert~\cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
the number of processes that access the data structure concurrently, which can be $p$ in the worst case.
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for any CAS-based implementation of a lock-free bag data structure, which provides operations
to insert an element or remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also applies to queues.
Although this might seem to settle the step complexity of lock-free queues, the lower bound
holds only if $c$ is $O(\log\log p)$ so it should be stated more precisely as
%At first glance, this would seem to settle the question of the step complexity of lock-free queues.
%However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
%Thus, the lower bound could be stated more precisely as 
an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

@@

Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored (in binary) in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations reach the root in batches and are linearized in the order they reach the root.
Since a batch can have up to $p$ operations, explicitly recording the list of operations composing a batch
or applying them one-by-one to the queue would be too costly.
Instead, we use a novel implicit representation of batches
of queue operations that allows us to quickly merge two batches from the children of a node,
and quickly access any  operation in a~batch.
A preliminary version of this work appeared in~\cite{Nad22}.
%\here{Maybe say a little more about techniques used in the implementation, if space permits}

>>>>

There has been a great deal of research in the past several decades on the design of shared queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g.,\cite{BBFRSW21}),
% ouroboros paper by Winter et al 2020 might be a source for queues used for mem management (in context of GPUs)
%packet processing \cite{DPDK}, 
synchronization \cite{KAE23}, %\here{is this a good citation for this?}
 and sharing resources or tasks.
We focus on shared queues that are \emph{linearizable} \cite{HW90}, meaning that operations
appear to take place atomically, and \emph{lock-free}, meaning that some operation on the queue
is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.
We study the \emph{amortized step complexity} per operation, which is measured by taking the maximum, over all possible
executions, of the number of steps in the execution divided by the total number of enqueue and dequeue operations in the execution.

@@

Many concurrent data structures that keep track of a set of elements also have an $\Omega(p)$ term in their step complexity, as observed in~\cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have an $\Omega(c)$ term in their step complexity, where $c$ represents contention,
the number of processes that access the data structure concurrently, which can be $p$ in the worst case.
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for any CAS-based implementation of a lock-free bag data structure, which provides operations
to insert an element or remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also applies to queues.
Although this might seem to settle the step complexity of lock-free queues, the lower bound
holds only if $c$ is $O(\log\log p)$ so it should be stated more precisely as
%At first glance, this would seem to settle the question of the step complexity of lock-free queues.
%However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
%Thus, the lower bound could be stated more precisely as 
an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

@@

Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored (in binary) in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.
\rev{See \cite{SBM24} and the references therein for a variety of approaches to
safe memory reclamation.}

Our queue uses a binary tree, called the \emph{\ordering\ tree}, where each process has its own leaf.
A process adds its operations to its leaf.
As in previous work (e.g., \cite{DBLP:conf/stoc/AfekDT95,DBLP:conf/fsttcs/JayantiP05}), operations are propagated from the leaves up to the root in a cooperative way that ensures wait-freedom
and avoids the CAS retry problem.
Operations reach the root in batches and are linearized in the order they reach the root.
Since a batch can have up to $p$ operations, explicitly recording the list of operations composing a batch
or applying them one-by-one to the queue would be too costly.
Instead, we use a novel implicit representation of batches
of queue operations that allows us to quickly merge two batches from the children of a node,
and quickly access any  operation in a~batch.
Preliminary versions of this work appeared in~\cite{Nad22,NR23}.
%\here{Maybe say a little more about techniques used in the implementation, if space permits}
